### 2011-11-01 paper.tex
paper.t

We define a minimal normal form of Tree-Adjoining Grammars (TAGs)
suited to biological sequence analysis,
as opposed to the linguistic representation elsewhere \cite{JoshiSchabes97}.
TAGs have previously been used in bioinformatics
to model pseudoknots and other RNA structures \cite{MatsuiEtAl2005,ChiangJoshiSearls2006}
and to model local duplications (Hickey and Blanchette, pers. comm.).

@@

A TAG is a tuple $\grammar=(\nodelabels,\terminals,\startsymbol,\rules,\weight)$ where
$\nodelabels$ is a finite set of {\em node labels},
$\terminals$ is a finite set of {\em terminals} (disjoint from $\nodelabels$),
$\startsymbol \in \nodelabels$ is a distinguished {\em start label},
$\rules$ is a finite set of {\em transformation rules}
and $\weight:\rules \to [0,\infty)$ is a {\em rule weight function}.

@@

\newcommand\tinyrhsA{\Tree[ .$\alpha$ [ .$C$ [ .$U$ $\epsilon$ ] [ .$\beta$ ] ] ]}
\newcommand\tinyrhsB{\Tree[ .$\alpha$ [ .$C$ [ .$\beta$ ] [ .$Y$ $\epsilon$ ] ] ]}
\newcommand\tinyrhsC{\Tree[ .$\alpha$ [ .$V$ $\epsilon$ ] [ .$C$ $\beta$ ] ]}
\newcommand\tinyrhsD{\Tree[ .$\alpha$ [ .$C$ $\beta$ ] [ .$X$ $\epsilon$ ] ]}
\newcommand\tinyrhsE{\Tree[ .$\alpha$ [ .$B$ [ .$C$ $\beta$ ] ] ]}
%\newcommand\tinyrhsF{\Tree[ .$\alpha$ $u$ [ .$C$ $v$ $\beta$ $x$ ] $y$ ]}
\newcommand\tinyrhsF{\abtree{C}}

@@

\newcommand\rhsA{\Tree[ .$\alpha$ [ .$U$ $\epsilon$ ] [ .$C$ $v$ $\beta$ $x$ ] $y$ ]}
\newcommand\rhsB{\Tree[ .$\alpha$ $u$ [ .$C$ $v$ $\beta$ $x$ ] [ .$Y$ $\epsilon$ ] ]}
\newcommand\rhsC{\Tree[ .$\alpha$ $u$ [ .$C$ [ .$V$ $\epsilon$ ] $\beta$ $x$ ] $y$ ]}
\newcommand\rhsD{\Tree[ .$\alpha$ $u$ [ .$C$ $v$ $\beta$ [ .$X$ $\epsilon$ ] ] $y$ ]}
\newcommand\rhsE{\rhsnoepsilon{B}{C}{u}{v}{x}{y}}
\newcommand\rhsF{\rhsnob{C}{u}{v}{x}{y}}

\newcommand\newickA{((\beta)A)\alpha \to (U,((v,\beta,x)C),y)\alpha}
\newcommand\newickB{((\beta)A)\alpha \to (u,((v,\beta,x)C),Y)\alpha}
\newcommand\newickC{((\beta)A)\alpha \to (u,((V,\beta,x)C),y)\alpha}
\newcommand\newickD{((\beta)A)\alpha \to (u,((v,\beta,X)C),y)\alpha}
\newcommand\newickE{((\beta)A)\alpha \to ((u,((v,\beta,x)C),y)B)\alpha}
\newcommand\newickF{((\beta)A)\alpha \to (u,((v,\beta,x)C),y)\alpha}

\newcommand\longnewickA{((\beta)A)\alpha \to ((\epsilon)U,((v,\beta,x)C),y)\alpha}

@@

We can further restrict the rules of \tabref{RuleTypes} to recover the basic repertoire of a context-free grammar.
For example, bifurcation rules ($A \to U\ C$)
may be interpreted as type-1 rules from \tabref{RuleTypes} (with $v=x=y=\epsilon$),
while emission rules ($A \to u\ C\ y$)
may be interpreted as type-6 rules (with $v=x=\epsilon$),

@@

The final column of \tabref{RuleTypes} shows a shorthand representation of each rule in Newick format,
probably the most widely-understood bioinformatics format for representing tree structures.
The initial tree, in this representation, is $((\epsilon)S)\epsilon$.
In the table, we have omitted the placeholder $\epsilon$'s at leaf nodes,
so that (for example) the rule
\[
\newickA
\]
should strictly be read as
\[
\longnewickA
\]

@@

Note the special case of a type-6 rule where all the terminal strings are empty,
$((\beta)A)\alpha \to ((\beta)C)\alpha$.
Such a rule is referred to as a {\em transition} and may be written more compactly as $A \to C$.

\begin{table}
\tiny
\begin{tabular}{m{.2in}m{.2in}m{.2in}m{1in}cm{1.2in}}
\centerline{Type} & \centerline{From} & & \centerline{To} & Newick representation & When $uvxy=\epsilon$:
\\ \hline
(1) &
\lhs & $\to$ & \rhsA & $\newickA$ & \tinyrhsA
\\ \hline
(2) &
\lhs & $\to$ & \rhsB & $\newickB$ & \tinyrhsB
\\ \hline
(3) &
\lhs & $\to$ & \rhsC & $\newickC$ & \tinyrhsC
\\ \hline
(4) &
\lhs & $\to$ & \rhsD & $\newickD$ & \tinyrhsD
\\ \hline
(5) &
\lhs & $\to$ & \rhsE & $\newickE$ & \tinyrhsE
\\ \hline
(6) &
\lhs & $\to$ & \rhsF & $\newickF$ & \tinyrhsF
\\ \hline
\end{tabular}
\normalsize
\caption{
\tablabel{RuleTypes}
Types of transformation rule (i.e. tree adjunction rule) used in this paper.
Here $\alpha,\beta$ represent any subtree;
$A \in \nodelabels$ is the source node label;
$B,C,U,V,X,Y \in (\nodelabels \cup \{\epsilon\})$ are the destination node labels;
$\epsilon$ is the empty string;
and
$u,v,x,y \in \terminals^\ast$ are (possibly empty) terminal strings.
The final column shows the destination tree when all the terminal strings $u,v,x,y$ are empty.
}
\end{table}

@@

Define some indicator functions to match output substrings to rules
\begin{eqnarray*}
\match(i,j,x) & = & \delta \left( \outsubseq{i}{j}\ =\ x \right) \\
\fmatch(i,x) & = & \match(i,i+|x|,x) \\
\rmatch(j,y) & = & \match(j-|y|,j,y) \\
\iumatch & = & \fmatch(i,u) \\
\jvmatch & = & \rmatch(j,v) \\
\kxmatch & = & \fmatch(k,x) \\
\lymatch & = & \rmatch(l,y)
\end{eqnarray*}

@@

The DP recursion relation is as follows (with $B,C,U,V,X,Y,u,v,x,y$ defined as in \tabref{RuleTypes})
\begin{eqnarray*}
\lefteqn{\m(i,j,k,l,A) =} \\
& &
\sum_{\rho \in \rulesubset{1}{A}} \sum_{m=i}^j W(\rho) \m(m,j-|v|,k+|x|,l-|y|,C) \mtip(i,m,U) \jvmatch \kxmatch \lymatch \\
& & +
\sum_{\rho \in \rulesubset{2}{A}} \sum_{n=k}^l W(\rho) \m(i+|u|,j-|v|,k+|x|,n,C) \mtip(n,l,Y) \iumatch \jvmatch \kxmatch \\
& & +
\sum_{\rho \in \rulesubset{3}{A}} \sum_{m=i}^j W(\rho) \m(i+|u|,m,k+|x|,l-|y|,C) \mtip(m,j,V) \iumatch \kxmatch \lymatch \\
& & +
\sum_{\rho \in \rulesubset{4}{A}} \sum_{n=k}^l W(\rho) \m(i+|u|,j-|v|,n,l-|y|,C) \mtip(k,n,X) \iumatch \jvmatch \lymatch \\
& & +
\sum_{\rho \in \rulesubset{5}{A}} \sum_{m=i}^j \sum_{n=k}^l W(\rho) \m(i+|u|,m,n,l-|y|,B) \m(m,j-|v|,k+|x|,n,C) \iumatch \jvmatch \kxmatch \lymatch \\
& & +
\sum_{\rho \in \rulesubset{6}{A}} W(\rho) \m(i+|u|,j-|v|,k+|x|,l-|y|,C)
 \iumatch \jvmatch \kxmatch \lymatch
\end{eqnarray*}

Note that for $\m$ to be exactly computable, we require that there are no {\em null cycles} in the grammar.
A null cycle is a series of transformations that, when applied consecutively to a given tree $\alpha$, yield the original tree $\alpha$ again,
or one that is trivially related to it (e.g. a tree that is identical to $\alpha$ after $\epsilon$-labeled nodes of degree $<3$ have been removed).
Null cycles frequently arise via consecutive transitions of the form $A \to B$ followed by $B \to A$,
but there are other possibilities too (for example, a type-1 bifurcation rule $((\beta)A)\alpha \to ((U,\beta)C)\alpha$
followed by two transitions $U \to \epsilon$ and $C \to A$).

A {\em Lexicalized TAG} (LTAG) has the property that one of $u,v,x,y$ is always a nonempty terminal string,
guaranteeing that it contains no null cycles,
and ensuring that every rule explicitly matches some output characters.
LTAGs are popular in linguistics, though the constraint is possibly too stringent to be helpful for bioinformatics grammars.

@@

The LTRs can be generated by type-6 rules (e.g. with $v=y=\epsilon$) followed either by a transition to the transposon interior,
or (optionally) a type-5 adjunction rule, which may be an easier event on which to impose a hint-requirement constraint.

@@

\noindent
\begin{tabular}{ll}
Symbol & Meaning \\
\hline
$\grammar$ & Grammar \\
$\nodelabels$ & Set of node labels \\
$\terminals$ & Set of terminals \\
$\rules$ & Set of transformation rules \\
$\rulesubset{n}{A}$ & Set of transformation rules of type $n$ (see \tabref{RuleTypes}) whose LHS is $((\beta)A)\alpha$ \\
$\alpha,\beta$ & Arbitrary ordered node-labeled trees \\
$\weight$ & Rule weight function \\
$\seqweight{\alpha}{\outseq}$ & Sum, over derivations of $\outseq$ from $\alpha$, of product over derivation rule weights \\
$\epsilon$ & The empty string \\
$\terminals^\ast$ & Set of strings over $\terminals$, including the empty string \\
$\outseq$ & Output sequence \\
$|\outseq|$ & Length of $\outseq$ \\
$i,j,k,l$ & Integer co-ordinates on $\outseq$, with $1 \leq i \leq j \leq k \leq l \leq |\outseq|$ \\
$m,n$ & Integer co-ordinates on $\outseq$, with $i \leq m \leq j$ and $k \leq n \leq l$ \\
$\outerlen, \innerlen$ & Integer lengths of $\outseq$-subsequences, with $0 \leq \innerlen = (k-j) \leq \outerlen = (l-i) \leq |\outseq|$ \\
$\outsubseq{i}{i}$ & The empty string \\
$\outsubseq{i}{j}$ & (For $j > i$) Substring of $\outseq$ from $i$ to $j-1$ inclusive \\
$\match(i,j,x)$ & Indicates if rule string $x$ matches output string $\outsubseq{i}{j}$ \\
$\fmatch(i,x)$ & Indicates if rule string $x$ matches output string $\outseq$, starting at position $i$ \\
$\rmatch(j,y)$ & Indicates if rule string $y$ matches output string $\outseq$, ending before position $j$ \\
$\iumatch,\jvmatch,\kxmatch,\lymatch$ & Rule string matchers; $\match_{ns}$ matches string $s$ ending/starting at position $n$ \\
$\m(i,j,k,l,A)$ & Inside weight for $\outsubseq{i}{j}$ (left) and $\outsubseq{k}{l}$ (right) rooted at $A$ \\
$\mtip(i,l,A)$ & Inside weight for $\outsubseq{i}{l}$ rooted at $A$ \\
$B,C,U,V,X,Y$ & Labels (or terminal strings) at nodes of the Newick tree $((U,((V,\beta,W)C),X)B)\alpha$ \\
$u,v,x,y$ & Special symbols for $U,V,X,Y$ when they are terminal strings \\
\end{tabular}

>>>>

We define a minimal normal form of Tree-Adjoining Grammars (TAGs)
suited to simplified presentation for biological sequence analysis,
departing somewhat from the usual linguistic representation \cite{JoshiSchabes97}.
TAGs have previously been used in bioinformatics
to model pseudoknots and other RNA structures \cite{MatsuiEtAl2005,ChiangJoshiSearls2006}
and to model local duplications (Hickey and Blanchette, pers. comm.).

@@

A TAG is a tuple $\grammar=(\nodelabels,\terminals,\startsymbol,\rules,\weight)$ where
$\nodelabels$ is a finite set of {\em node labels} (or {\em nonterminals}),
$\terminals$ is a finite set of {\em terminals} (disjoint from $\nodelabels$),
$\startsymbol \in \nodelabels$ is a distinguished {\em start label},
$\rules$ is a finite set of {\em transformation rules}
and $\weight:\rules \to [0,\infty)$ is a {\em rule weight function}.

@@

\newcommand\tinyrhsBif{\Tree[ .$\alpha$ [ .$V$ $\epsilon$ ] [ .$\beta$ ] [ .$X$ $\epsilon$ ] ]}

\newcommand\tinyrhsA{\Tree[ .$\alpha$ [ .$C$ [ .$U$ $\epsilon$ ] [ .$\beta$ ] ] ]}
\newcommand\tinyrhsB{\Tree[ .$\alpha$ [ .$C$ [ .$\beta$ ] [ .$Y$ $\epsilon$ ] ] ]}
\newcommand\tinyrhsC{\Tree[ .$\alpha$ [ .$V$ $\epsilon$ ] [ .$C$ $\beta$ ] ]}
\newcommand\tinyrhsD{\Tree[ .$\alpha$ [ .$C$ $\beta$ ] [ .$X$ $\epsilon$ ] ]}

\newcommand\tinyrhsE{\Tree[ .$\alpha$ [ .$B$ [ .$C$ $\beta$ ] ] ]}
%\newcommand\tinyrhsF{\Tree[ .$\alpha$ $u$ [ .$C$ $v$ $\beta$ $x$ ] $y$ ]}
\newcommand\tinyrhsF{\abtree{C}}

@@

\newcommand\rhsBif{\Tree[ .$\alpha$ [ .$V$ $\epsilon$ ] $\beta$ [ .$X$ $\epsilon$ ] ]}

\newcommand\rhsA{\Tree[ .$\alpha$ [ .$U$ $\epsilon$ ] [ .$C$ $v$ $\beta$ $x$ ] $y$ ]}
\newcommand\rhsB{\Tree[ .$\alpha$ $u$ [ .$C$ $v$ $\beta$ $x$ ] [ .$Y$ $\epsilon$ ] ]}
\newcommand\rhsC{\Tree[ .$\alpha$ $u$ [ .$C$ [ .$V$ $\epsilon$ ] $\beta$ $x$ ] $y$ ]}
\newcommand\rhsD{\Tree[ .$\alpha$ $u$ [ .$C$ $v$ $\beta$ [ .$X$ $\epsilon$ ] ] $y$ ]}

%\newcommand\rhsE{\rhsnoepsilon{B}{C}{u}{v}{x}{y}}
\newcommand\rhsE{\tinyrhsE}
\newcommand\rhsF{\rhsnob{C}{u}{v}{x}{y}}

\newcommand\newickBif{((\beta)A)\alpha \to (V,\beta,X)\alpha}

\newcommand\newickA{((\beta)A)\alpha \to (U,((v,\beta,x)C),y)\alpha}
\newcommand\newickB{((\beta)A)\alpha \to (u,((v,\beta,x)C),Y)\alpha}
\newcommand\newickC{((\beta)A)\alpha \to (u,((V,\beta,x)C),y)\alpha}
\newcommand\newickD{((\beta)A)\alpha \to (u,((v,\beta,X)C),y)\alpha}

%\newcommand\newickE{((\beta)A)\alpha \to ((u,((v,\beta,x)C),y)B)\alpha}
\newcommand\newickE{((\beta)A)\alpha \to (((\beta)C)B)\alpha}
\newcommand\newickF{((\beta)A)\alpha \to (u,((v,\beta,x)C),y)\alpha}

\newcommand\longnewickBif{((\beta)A)\alpha \to ((\epsilon)V,\beta,(\epsilon)X)\alpha}

@@

We can further restrict the rules of \tabref{RuleTypes} to recover the basic repertoire of a context-free grammar,
or the weighted (stochastic) version, the SCFG.
For example, SCFG rules of the form $A \to V\ X$, where $A,V,X$ are nonterminals,
may be interpreted as bifurcation rules in \tabref{RuleTypes},
while SCFG rules of the form $A \to u\ C\ y$, where $u,y$ are terminal strings and $C$ is either a nonterminal or the empty string,
may be interpreted as emission rules in \tabref{RuleTypes}.

@@

The final column of \tabref{RuleTypes} shows a shorthand representation of each rule in Newick format,
probably the most widely-understood bioinformatics format for representing tree structures.
The initial tree, in this representation, is $((\epsilon)S)\epsilon$.
In the table, we have omitted the placeholder $\epsilon$'s at leaf nodes,
so that (for example) the rule
\[
\newickBif
\]
should strictly be read as
\[
\longnewickBif
\]

@@

Note the special case of an emission rule where all the terminal strings are empty,
$((\beta)A)\alpha \to ((\beta)C)\alpha$.
Such a rule is referred to as a {\em transition} and may be written more compactly as $A \to C$.

\begin{table}
\tiny
\begin{tabular}{m{.4in}m{.2in}m{.2in}m{1in}c}
\centerline{Type} & \centerline{From} & & \centerline{To} & Newick representation
\\ \hline
(BIFURC) &
\lhs & $\to$ & \rhsBif & $\newickBif$
\\ \hline
(ADJOIN) &
\lhs & $\to$ & \rhsE & $\newickE$
\\ \hline
(EMIT) &
\lhs & $\to$ & \rhsF & $\newickF$
\\ \hline
\end{tabular}
\normalsize
\caption{
\tablabel{RuleTypes}
Types of transformation rule (i.e. tree adjunction rule) used in this paper.
Here $\alpha,\beta$ represent any subtree;
$A \in \nodelabels$ is the source node label;
$B,C,V,X \in (\nodelabels \cup \{\epsilon\})$ are the destination node labels;
$\epsilon$ is the empty string;
and
$u,v,x,y \in \terminals^\ast$ are (possibly empty) terminal strings.
The final column shows the destination tree when all the terminal strings $u,v,x,y$ are empty.
}
\end{table}

@@

Define some indicator functions to match output substrings to emit rules
\begin{eqnarray*}
\match(i,j,x) & = & \delta \left( \outsubseq{i}{j}\ =\ x \right) \\
\fmatch(i,x) & = & \match(i,i+|x|,x) \\
\rmatch(j,y) & = & \match(j-|y|,j,y) \\
\iumatch & = & \fmatch(i,u) \\
\jvmatch & = & \rmatch(j,v) \\
\kxmatch & = & \fmatch(k,x) \\
\lymatch & = & \rmatch(l,y)
\end{eqnarray*}

@@

The DP recursion relation is as follows (with $B,C,V,X,u,v,x,y$ defined as in \tabref{RuleTypes})
\begin{eqnarray*}
\lefteqn{\m(i,j,k,l,A) =} \\
& &
\sum_{\rho \in \rulesubset{BIF}{A}} \sum_{m=i}^j W(\rho) \mtip(i,j,V) \mtip(k,l,X) \\
& & +
\sum_{\rho \in \rulesubset{ADJ}{A}} \sum_{m=i}^j \sum_{n=k}^l W(\rho) \m(i,m,n,l,B) \m(m,j,k,n,C) \\
& & +
\sum_{\rho \in \rulesubset{EMT}{A}} W(\rho) \m(i+|u|,j-|v|,k+|x|,l-|y|,C) \iumatch \jvmatch \kxmatch \lymatch
\end{eqnarray*}

Note that for $\m$ to be exactly computable, we require that there are no {\em null cycles} in the grammar.
A null cycle is a series of transformations that, when applied consecutively to a given tree $\alpha$, yield the original tree $\alpha$ again,
or one that is trivially related to it (e.g. a tree that is identical to $\alpha$ after $\epsilon$-labeled nodes of degree $<3$ have been removed).
Null cycles frequently arise via consecutive transitions of the form $A \to B$ followed by $B \to A$,
but there are other possibilities too (for example, a bifurcation rule $((\beta)A)\alpha \to (V,\beta,X)\alpha$
followed by two transitions $V \to \epsilon$ and $X \to A$).

@@

The LTRs can be generated by emission rules (e.g. with $v=y=\epsilon$) followed either by a transition to the transposon interior,
or (optionally) an adjunction rule, which may be an easier event on which to impose a hint-requirement constraint.

@@

\noindent
\begin{tabular}{ll}
Symbol & Meaning \\
\hline
$\grammar$ & Grammar \\
$\nodelabels$ & Set of node labels (nonterminals) \\
$\terminals$ & Set of terminals \\
$\rules$ & Set of transformation rules \\
$\rulesubset{T}{A}$ & Set of type-$T$ transformation rules (see \tabref{RuleTypes}) with $((\beta)A)\alpha$ on LHS \\
$\alpha,\beta$ & Arbitrary ordered node-labeled trees \\
$\weight$ & Rule weight function \\
$\seqweight{\alpha}{\outseq}$ & Sum, over derivations of $\outseq$ from $\alpha$, of product over derivation rule weights \\
$\epsilon$ & The empty string \\
$\terminals^\ast$ & Set of strings over $\terminals$, including the empty string \\
$\outseq$ & Output sequence \\
$|\outseq|$ & Length of $\outseq$ \\
$i,j,k,l$ & Integer co-ordinates on $\outseq$, with $1 \leq i \leq j \leq k \leq l \leq |\outseq|$ \\
$m,n$ & Integer co-ordinates on $\outseq$, with $i \leq m \leq j$ and $k \leq n \leq l$ \\
$\outerlen, \innerlen$ & Integer lengths of $\outseq$-subsequences, with $0 \leq \innerlen = (k-j) \leq \outerlen = (l-i) \leq |\outseq|$ \\
$\outsubseq{i}{i}$ & The empty string \\
$\outsubseq{i}{j}$ & (For $j > i$) Substring of $\outseq$ from $i$ to $j-1$ inclusive \\
$\match(i,j,x)$ & Indicates if rule string $x$ matches output string $\outsubseq{i}{j}$ \\
$\fmatch(i,x)$ & Indicates if rule string $x$ matches output string $\outseq$, starting at position $i$ \\
$\rmatch(j,y)$ & Indicates if rule string $y$ matches output string $\outseq$, ending before position $j$ \\
$\iumatch,\jvmatch,\kxmatch,\lymatch$ & Rule string matchers; $\match_{ns}$ matches string $s$ ending/starting at position $n$ \\
$\m(i,j,k,l,A)$ & Inside weight for $\outsubseq{i}{j}$ (left) and $\outsubseq{k}{l}$ (right) rooted at $A$ \\
$\mtip(i,l,A)$ & Inside weight for $\outsubseq{i}{l}$ rooted at $A$ \\
$B,C,U,V,X,Y$ & Labels (nonterminal symbols, terminal strings, or $\epsilon$) at nodes of the Newick tree $((U,((V,\beta,W)C),X)B)\alpha$ \\
$u,v,x,y$ & Special symbols for $U,V,X,Y$ when they are terminal strings \\
\end{tabular}
